Your time series is (probably) too short

### 2017-07-07 writing/manuscript_draft.Rmd
added citations and modified code for figure 1

# Abstract
Long-term census data is necessary to better understand ecological processes and make management decisions. However, census data is often expensive, requiring a lot of people-hours and equipment. Little work has addressed the question, when is a time series of census data long enough to address a question of interest? Here, we explore two approaches to address this question: through simulations and through an empirical approach. I specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, I examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability. In addition, we examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. <!--Here, We stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.

@@

Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010]. long-term data is necessary to both understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "long-term", most are still short (i.e. less than X years). *I could have statistic on how long is considered "long-term" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time (citations). There are, of course, notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others (citations).

@@

For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker). Instead of using classic statistical tests (e.g. t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987]. Simple models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then, time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker). In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series [@Gerber1999]. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

@@

```{r,echo=F,eval=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power. (c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
source("../scripts/theoretical-models/calculate_power.R")
source("../scripts/theoretical-models/simulate_pop.R")

  years_to_sample=seq(2,100,by=1)
  power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=50)#should use 300+ trials

  simulations = replicate(n=30,simulate_pop(phi = 0.6,r=0.2,sigma = .5,max.time=40))
  min_time_needed = years_to_sample[tail(which(power<0.8),1)] + 1

  # Plotting commands
  par(mfrow=c(2,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
    matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
      mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'population size' ,side = 2,cex = 1.2,line=4)
      mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

    # Plot versus trend strength
      r_values = seq(1,3,by=1) #by=0.2
      sigma_values=seq(1,7,by=1) #by=0.2

      parameter_combinations = expand.grid(r_values,5)
      parameter_combinations$min_time_required = NA
      names(parameter_combinations)= c('r','sigma','min_time_required')

      for (i in 1:nrow(parameter_combinations)){
        years_to_sample=seq(2,50,by=1)
        y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=100)
        if (length(tail(which(y<0.8),1))>0){
        parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
        }
        #print(i)
      }

@@

      # Plot versus sigma values
      parameter_combinations = expand.grid(1.5,sigma_values)
      parameter_combinations$min_time_required = NA
      names(parameter_combinations)= c('r','sigma','min_time_required')

      for (i in 1:nrow(parameter_combinations)){
        years_to_sample=seq(2,50,by=1)
        y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=100)
        if (length(tail(which(y<0.8),1))>0){
        parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
        }
        #print(i)
      }

      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
      mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(d)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
```

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation). <!--In Figure \ref{fig:empirical_approach_example}b, statistical power increases with more time sampled until our desired power of 0.8 at 16 years. Therefore, 16 years would be our estimate for $T_{min}$.-->

We also examine how minimum time series calculations changed depending on the question of interest. In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig. \ref{fig:growth_rate}), the variability in population size, and future extinction risk. Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.

<!--Question 2: How sensitive are IUCN assessments to time series trends?
The IUCN sets threat statuses for a variety of species. Importaly, they have a threshold of 10 years or 3 generation. Therefore, we subsampled all possible 10-year, or greater, time series for each population. For example, species X has been sampled for 40 years. This would lead to XX possible 10-year time series. We then determined what percentage of these 10 year, or greater, time series would lead to the wrong conculsions compared to the full time series.

Question 3: How long does a population need to be sampled to capture the variability in population size?

Populations vary from year-to-year for several reasons including-->

# Results
<!--
- summary of findings
- theoretical approach
  - example plot
- empirical approach
  - plots: distribution of min time, what min time correlations with
  - walk through one empirical example (maybe in appendix)
-->

@@

## Questions in the appendix

Describe some of the results that we include in the appendix

@@

<!-- 
- Findings and how they relate to past work
- Extensions to other systems. Useful to use similar species
- Biological variables not good predictors, but we also used all long-lived vertebrate species
- Other questions to ask and results in appendix

- Future recommendations to others and future work
- implications
- conclusion paragraph

The IUCN has an 10 year/3 generation rule... If we apply this rule to the species in our database we see that approximately `r 100*sum(pop_info$min_time_for_power>pop_info$IUCN)/nrow(pop_info)`% of populations need a longer minimum time series than that required by IUCN.
-->

We explored two different approaches to estimate the minimum time series length required to address a particular question of interest. We specifically asked, what is the minimum time series length required to determine long-term changes in abundance using linear regression? The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987, Johnson2015]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].

Here, we focus on a empirical approach to estimate the minimum time series length requried to assess changes in abundance over time. We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression. Statistical power is important as it provides on information as to the neccessary samples required to dertermine a significant trend [@Legg2006]. We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig. \ref{fig:min_time_dist}). This timeframe is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. In line with theoretical predictions (citations), we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig. \ref{fig:correlates}). Against our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (cite figure or supp mat). We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. Our result could have result of at least two effects. First, the database we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreases, is only worried about population surveys. Therefore, biological variables may be more important if the question itself was different. ANOTHER SENTENCE HERE

There has also been interest in determining the minimum time series length in the bird community [@Hatch2003, @Bart2004, @Seavy2007]. @Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power. He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design. Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations. They found that only about 42% of populations in the Breeding Bird Survey met their criteria. @Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for precision based metrics ..

An important related question, is the optimal allocation of sampling effort in space versus time (citations). In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (cite figure here). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term census of the collared pika (\emph{Ochotona collaris}) found in the Yukon. They estimated long-term growth rates among three subpopulations over a 10-year period. They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby, may be difficult. Therefore, they argued more sampling should be conducted spatially than temporally [@Morrison2008].

## Other questions addressed

In the appendix we asked two additional questions to make estimates of $T_{min}$. First, we asked how many years of sampling are neccessary to estimate a geometric growth rate?

@@

Our work has some limitations in determining the minimum time series length required. First, $T_{min}$ depends strongly on the question of interest. For our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another. In addition, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (see appendix). <!-- include sentence on how T_min depends on these parameters --> In an ideal world, we would build a specific population model parameterized for each population of interest. Then, model simulations could be used to estimate the minimum time series reuqired to address a specific question of interest. Clearly, this is not always practical, especially if running analyses for a wide array of species as we do here. In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression. Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required. We show that to identify long-term changes in abundance, 10-15 years of continuos monitoring are required (Fig. \ref{fig:min_time_dist}). In line with theoretical predictions, we also show that $T_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig. \ref{fig:correlates}). Our work implies that for many populations, census data less than 10-15 years are probably not reliable. This stresses the importance of long-term monitoring programs. <!--(where do IUCN recommendations come from?)--> From both a scientific and management perspective estimates of $T_{min}$ are important. If a time series is too short, we are likely to make incorrect conclusions about long-term trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Future work can examine other species, with a wider range of life history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address other questions.

@@

In the supporting material, we provide an expanded methods sections, additional figures, and additional analyses of other time series questions. All code and data can be found at www.github.com/erwhite1

@@

ERW was partially supported by a National Science Foundation Graduate Fellowship. We would like to thank members of the Ecological Theory group at the University of California, Davis for their insight. We would also like to thank X anonymous reviewers for their helpful comments.

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
time. The line is the best fit line from linear regresssion. (b)
Statistical power for different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

>>>>

# Abstract
Long-term census data is necessary to better understand ecological processes and make management decisions. However, census data is often expensive, requiring a lot of people-hours and equipment. Little work has addressed the question, when is a time series of census data long enough to address a question of interest? Here, we explore two approaches to address this question: through simulations and through an empirical approach. We specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, We examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability. In addition, we examine 868 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. <!--Here, We stress the need for verifying a study has enough statistical power to accurately determine long-term changes in population abundance.--> Most studies, especially those less than 10-15 years, are probably underpowered and potentially misleading.

@@

Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017]. Long-term data is necessary to both understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "long-term", most are still short. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017]]. There are, of course, notable exceptions where long-term monitoring has occured, for example the continuous plankton recorder [@Giron-Nava2017].

@@

For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power [@Bolker2008]. Instead of using classic statistical tests (e.g. t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987, Bolker2008]. Simple models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then, time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question [@Bolker2008]. In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series [@Gerber1999]. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

@@

```{r,echo=F,eval=T,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power. (c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
source("../scripts/theoretical-models/calculate_power.R")
source("../scripts/theoretical-models/simulate_pop.R")

  # years_to_sample=seq(2,100,by=1)
  # power=lapply(X=years_to_sample,FUN = calculate_power,phi = 0.6,r=0.2,sigma = .5,trials=100)#should use 300+ trials
  # 
  # simulations = replicate(n=30,simulate_pop(phi = 0.6,r=0.2,sigma = .5,max.time=100))
  # min_time_needed = years_to_sample[tail(which(power<0.8),1)] + 1
  # 
  # save(simulations,years_to_sample,power,min_time_needed,file='../analysis-outputs/example_simulations_fig1.Rdata')
  load('../analysis-outputs/example_simulations_fig1.Rdata')

  #load command
  # Plotting commands
  par(mfrow=c(2,2),oma=c(1,2,0,0.5),mar=c(4,5,0.5,0.5))
    matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
      mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'population size' ,side = 2,cex = 1.2,line=4)
      mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

    #   
    # # Plot versus trend strength
    #   r_values = seq(1,3,by=0.1) #by=0.2
    #   sigma_values=seq(1,7,by=0.3) #by=0.2
    # 
    #   parameter_combinations = expand.grid(r_values,5)
    #   parameter_combinations$min_time_required = NA
    #   names(parameter_combinations)= c('r','sigma','min_time_required')
    #   
    #   for (i in 1:nrow(parameter_combinations)){
    #     years_to_sample=seq(2,50,by=1)
    #     y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=150)
    #     if (length(tail(which(y<0.8),1))>0){
    #     parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
    #     }
    #     #print(i)
    #   }
    #   
    #   save(parameter_combinations,file='../analysis-outputs/min_time_vs_trend_strength.Rdata')
      load('../analysis-outputs/min_time_vs_trend_strength.Rdata')

@@

      # 
      # 
      # 
      # # Plot versus sigma values
      # parameter_combinations = expand.grid(1.5,sigma_values)
      # parameter_combinations$min_time_required = NA
      # names(parameter_combinations)= c('r','sigma','min_time_required')
      # 
      # for (i in 1:nrow(parameter_combinations)){
      #   years_to_sample=seq(2,50,by=1)
      #   y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=parameter_combinations$r[i],sigma = parameter_combinations$sigma[i],trials=150)
      #   if (length(tail(which(y<0.8),1))>0){
      #   parameter_combinations$min_time_required[i] = years_to_sample[tail(which(y<0.8),1)] + 1
      #   }
      #   #print(i)
      # }

      #save(parameter_combinations,file='../analysis-outputs/min_time_vs_pop_var.Rdata')
      load('../analysis-outputs/min_time_vs_pop_var.Rdata')
      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
      mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'minimum time required' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(d)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
```

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999; @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1 - \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation).

<!--
## Questions addressed in the appendix

We also examine how minimum time series calculations changed depending on the question of interest. In the supplementary material, we use similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig. \ref{fig:growth_rate}). We also explore how the method can be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}. More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. Lastly, we determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here.
-->

# Results

@@

## Questions in the supplmentary material

In the supplmentary material, we explore several additional questions related to statistical power and estimating $T_{min}$. First, we showed that other metrics besides estimating the slope coefficient from linear regression could be used. More specifically, we show how the same approach desribed here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs. \ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also explored how the simulation approach can be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}). More specifically, we determine the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. We found that over 27 years of continous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}. Lastly, we determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig. \ref{fig:gam_example}). We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig. \ref{fig:min_time_dist_gam}). We expand on these results in the supplementary material.

@@

We explored two different approaches to estimate the minimum time series length required to address a particular question of interest. We specifically asked, what is the minimum time series length required to determine long-term changes in abundance using linear regression? The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].

Here, we focus on an empirical approach to estimate the minimum time series length requried to assess changes in abundance over time. We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression. Statistical power is important as it provides on information as to the neccessary samples required to determine a significant trend [@Legg2006]. We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were neccessary (Fig. \ref{fig:min_time_dist}). This timeframe is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig. \ref{fig:biological_correlates}). We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. Our result could have been a result of at least two effects. First, the database we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time. Therefore, life-history characteristics may be more important for other questions (e.g. extinction risk) where estimates minimum time required are needed.

There has been interest in determining the minimum time series length in the bird research community [@Hatch2003; @Bart2004; @Seavy2007]. For example, @Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power. He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design. Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations. They found that only about 42% of populations in the Breeding Bird Survey met their criteria. @Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision. For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.

An important related question, is the optimal allocation of sampling effort in space versus time. In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig. \ref{fig:correlates}). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term census of the collared pika (\emph{Ochotona collaris}) found in the Yukon. They estimated long-term growth rates among three subpopulations over a 10-year period. They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.

@@

Our work has some limitations in determining the minimum time series length required. First, $T_{min}$ is particular to the specific question of interest. For example, in the supplementary material, we used generalized additive models to detect population trends and came up with a similar, but quantivaely different value for $T_{min}$ (Fig. \ref{fig:min_time_dist_gam}). An additional complication is that for our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another. Further, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (Fig. \ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a specific population model parameterized for each population of interest. Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest. Clearly, this is not always practical, especially if conducting analyses for a wide array of species as we do here. In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression (Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to determine population trends. This goes beyond previous work that either focused on theoretical investigations or a limited number of species. We show that to identify long-term changes in abundance, 10-15 years of continuos monitoring are required (Fig. \ref{fig:min_time_dist}). In line with theoretical predictions [@Gerrodette1987], we also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e. rate of increase), variability in population size, and the temporal autocorrelation (Fig. \ref{fig:correlates}). Our work implies that for many populations, census data less than 10-15 years are probably not reliable. This stresses the importance of long-term monitoring programs. From both a scientific and management perspective estimates of $T_{min}$ are important. If a time series is too short we lack statistical power to reliably detect long-term population trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Future work should examine other species, with a wider range of life history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.

@@

In the supporting material, we provide an expanded methods sections, additional figures, minimum time calculations for determining exponetial growth, simulations with a more complicated population model, minimum time calculations for determining long-term growth rates, and the use of generalized additive models to identify population trends. All code and data can be found at www.github.com/erwhite1

@@

ERW was partially supported by a National Science Foundation Graduate Fellowship. We would like to thank members of the Ecological Theory group at the University of California, Davis for their insight. We would also like to thank T. Dallas and E. Malcolm-White<!--X anonymous reviewers -->for their helpful comments.

@@

\begin{figure}[htbp]
\centering
\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
time. The line is the best fit line from linear regresssion. (b)
Statistical power for different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{Output of generalized linear model with a Poisson error
structure for predicting the minimum time required with explanatory
variables of the absolute value of the slope coefficient (or trend
strength), temporal autocorrelation, and variability in population
size.\label{fig:poisson_model}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{(a) Minimum time required to estimate change in abundance for
species class, (b) long-term trend (estimated slope coefficient) by
species class, (c) interannual variability in population size by species
class, and (d) temporal autocorrelation by species
class.\label{fig:class}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{Minimum time required to assess long-term trends in abundance
for values of statistical significance (\(\alpha\)) and power
(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in log(abundance) given power of
0.8.\label{fig:min_time_dist_log_pop}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{Statistical power for different length of time series
simulations for a lemon shark population in Bimini,
Bahamas.\label{fig:shark_example}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{Example of calculating minimum time required for growth rate
estimation. (a) European herring gull (\emph{Larus argentatus}) scaled
density over time, (b) mean and standard deviation of growth rate for
subsamples of entire time series, and (c) the percent error between mean
estimated growth rate and the true long-term growth rate. The vertical
bar denotes the minimum time required to estimate growth rate within
20\% error.\label{fig:growth_rate}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{Histogram of the minimum time required in order to estimate the
long-term growth rate within 20\%
error.\label{fig:min_time_growth_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
corresponding fitted GAM model in red and (b) statistical power as a
function of the number of years sampled. The horizontal ine at 0.8
indicates the minimum threshold for statistical power and the vertical
line denotes the minimum time required to achieve 0.8 statistical
power.\label{fig:gam_example}}
\end{figure}

\begin{figure}[htbp]
\centering
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance according to a GAM
model given statistical power of 0.8. The smoothing parameter was set to
3 for each population.\label{fig:min_time_dist_gam}}
\end{figure}
